On the properties of discs around accreting brown dwarfs

### 2010-07-13 BDD_submit/paper.tex
Some spelling mistakes fixed. Thanks Dave Rundle
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2368 25464f21-6099-402c-b30a-640f1d40d38b

There is now strong evidence that as brown dwarfs (BDs) form they pass
through a Classical T~Tauri star (CTTS) phase, during which they
possess a flared, dusty circumstellar disc from which they are
actively accreting material
\citep{jayawardhana_2003,mohanty_2004}. The accretion is thought to
proceed via a magnetically-controlled funnel flow mechanism in which
material from a truncated inner disc boundary falls onto the surface
of the star along magnetic field lines
\citep{camenzind_1990,koenigl_1991,muzerolle_2003,mohanty_2008}.

The interpretation of the spectral energy distributions (SEDs) of
pre-main-sequence (PMS) stars involves trying to distinguish the
various contributions to the continuum from the hot spots at the base
of the accretion flow, the photospheric flux, and the near-IR flux
from the dusty inner disc. The complexity of the interplay between
these contributions is exacerbated by other, geometrical, effects such
as the inclination (which changes the projected area of the inner disc
wall visible to the observer) and the outer disc structure (which may
obscure the inner disc for high inclinations)
\citep{walker_2004,tannirkulam_2007}.

In surveys, particularly those attempting to discern the PMS disc
fraction this disentangling usually takes place using broad-band
photometric measures and cuts in colour-colour or colour-magnitude
space \citep{luhman_2005b,luhman_2008,gutermuth_2008,monin_2010}.  The
important quantity here is the separation in wavelength between the
emission peaks for the stellar and thermal disc components, as it is
these components which must be isolated. For BD disc (BDD hereafter)
systems it is likely that difficulties detecting the disc will be
exacerbated by the lower temperatures of the photosphere and therefore
the smaller separation in wavelength from the thermal disc emission
component, compared to CTTS systems. Therefore, to derive disc
fractions for BD populations we require detailed comparison models
with which to guide disc candidate selection.

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be independent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}. At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photospheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars \citep[although this boundary may well be
spectral type dependent][]{navascues_2003}. Such objects should show
doppler-broadened profiles, but the line wings will be weak, and the
presence of underlying H$\alpha$ absorption may become dominant. Thus
there may well be a population of low-accretion-rate CTTS which are
current missing from the surveys.

@@

In this Section we detail the physical model adopted and assumptions
made (Section \ref{physics}), then explain key elements of the
radiative transfer code (Section \ref{torus}). Then we discuss the
derived values, such as broadband photometric magnitudes and colours
(in Section \ref{derived}). Some internal consistency checks are
given in Appendix \ref{consistency}.

@@

In order to model an accreting BDD system we must first model the
underlying photospheric flux. We have adopted a BD stellar interior
and atmospheric model grid and have then constructed the total
photospheric flux for any input value of stellar age and mass by
interpolating for surface gravity ($\log g$), effective temperature
($T_{\rm eff}$), radius ($R_*/R_{\odot}$) and luminosity
($L_*/L_{\odot}$). These values were then used to interpolate
atmospheric spectra for flux (ergs s$^{-1}$cm$^{-2} {\rm \AA}^{-1}$)
from 1200 to 2$\times$10$^7$ \AA. The spectra were subsequently
resampled onto 200 logarithmically spaced points. Careful inspection
ensured that no spectral features were removed during resampling. The
stellar interior models used for this study are the `DUSTY00' models
of \cite{chabrier_2000} combined with the `AMES-Dusty', atmospheric
models of \cite{chabrier_2000}, which are all available
online\footnote{http://perso.ens-lyon.fr/france.allard/}. For our
$T_{\rm eff}$ range of $\approx $3000\,K$<T_{\rm eff}<$1600\,K the
AMES-Dusty atmospheres are the most applicable (2700\,K$>T_{\rm
  eff}>$1700\,K). We did try including dynamic application of
atmospheres based on the derived $T_{\rm eff}$, i.e. using AMES-Cond
for $T_{\rm eff}<$1700\,K, but this resulted in large discontinuities
between the model atmospheres and resulting spectra. Additionally, for
the higher temperatures in our range \cite{martin_2000} have found the
NextGen models to be more applicable for $\sim T_{\rm
  eff}>2300$K. However, as with the lower boundary this only applies
to the edge of our temperature range and we feel it is better to use a
consistent set of atmosphere and interior models across our
grid. Since this only affects stars at the very edge of our
temperature range, i.e. for the oldest and lowest mass objects (for
the AMES-Cond case), we have adopted the AMES-Dusty models throughout.

@@

Finally, a blackbody flux distribution is generated at $T_{\rm acc}$
and added onto the intrinsic stellar photospheric flux. In general one
would expect this to be an overestimate of the accretion flux, as for
BDs large convective zones are expected on the stellar surface, and
some of the accretion energy may act to further drive these convective
currents, meaning flux is lost. It is worth noting however that
observationally UV excesses are often used to recreate and then
subtract an assumed accretion flux using a blackbody flux curve, which
is essentially the reverse of this method.

@@

In this study we assume that accretion from the central star occurs
along magnetically channelled columns from the inner disc boundary.
For CTTS stars, \cite{bouvier_2007} show that the magnetic truncation
radius ($R_{\rm mag}$) is less than the co-rotation radius ($R_{\rm
  co}$), where the angular Keplerian velocity of the disc is equal to
the surface angular velocity of the central star.  Calculations of the
magnetic truncation radius depend on derivations of the surface
magnetic field \citep{koenigl_1991}. This is currently unavailable for
BD stars, due to increased molecular species obscuring the Zeeman
splitting signatures that are normally used to derive stellar surface
magnetic fields. Therefore, for our model grid we have adopted an
initial inner disc radius as the co-rotation radius,
\begin{equation}
{R_{\rm inner}=\left({GM_*\tau^2\over{4\pi ^2}}\right)^{1\over{3}}},
\label{inner_eq}
\end{equation}
where $\tau$ is the stellar rotation period and $R_{\rm inner}$ is the
inner radius. This is effectively adopting a disc-locking mechanism
(without associated angular momentum loss), as for disc-locked stars,
$R_{\rm mag}\approx R_{\rm co}$,
\citep{koenigl_1991,shu_1994}. Therefore, for our
model simulations, this inner edge radius is dependent on, and derived
from, the adopted value of the rotational period for the central star,
as well as being weakly dependent on the stellar mass. As discussed in
Section \ref{intro}, the inner disc can be cleared through a number of
mechanisms, including binarity or giant planet formation,
photoevaporation or photoionisation of the disc and dust grain growth
or settling. For BDD systems where a disc is modeled a treatment of
dust sublimation is included (discussed in Section \ref{dust_edge}).
However, the effects of binarity or giant planet formation are
neglected. Further to the stellar mass and period required prior to
calculation of the inner disc radius, we require a disc mass (in
stellar masses).

@@

We have used the {\sc torus} radiative transfer code which was
originally described by \cite{harries_2000} and was subsequently
updated by \cite{harries_2004} and \cite{kurosawa_2004}. {\sc torus}
  uses the method of \cite{lucy_1999} solve radiative equilibrium on
  an AMR mesh. The code also self-consistently solves the
  equation of vertical hydrostatic equilibrium and dust sublimation
  for the disc \citep[described in][]{tannirkulam_2007}. The {\sc
    torus} code has been extensively benchmarked \citep{pinte_2009}.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range \citep{dullemond_2001}. The
circumstellar disc receives direct photospheric radiation and is
strongly heated. This leads to an increased scaleheight, or `puffing
up' of the inner disc rim, and subsequent shadowing of the region
immediately beyond it \citep{dullemond_2001}. The density-dependent
nature of the dust sublimation temperature \citep{pollack_1994} may
also play a role in shaping the inner rim
\citep{isella_2005,tannirkulam_2007}\footnote{Although we recognise
  that the inner hole in protostellar discs may be the result of giant
  planet formation or photoevaporation \citep{dahm_2009,najita_2007},
  these two mechanisms should result in a negligible mass-accretion
  rate \citep{dahm_2009} and therefore would represent a distinct
  evolutionary state of brown dwarf disc (BDD) systems to that of the
  CTTS-like phase we are considering here.}.

@@

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a swift
convergence of the sublimated rim. The dust sublimation proceeds as
follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessitates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations. For a more
sophisticated, and chemically more complex, study of dust sublimation,
albeit for higher mass objects see \cite{kama_2009}.

For our model grid we have adopted a gas-to-dust ratio, $\epsilon$, of
100, and the gas population is assumed to be essentially static with a
zero optical depth.

@@

Once the radiative transfer code was completed simulated SEDs were
generated. These SEDs can be generated for any distance and for any
system-observer inclination. For our models we have set the distance,
to 10\,pc to create absolute flux SEDs and selected ten inclinations
equi-spaced in $\cos i$ (0, 27, 39, 48, 56, 64, 71, 77, 84 and
90$^{\circ}$). A further useful feature of the {\sc torus} code is
that the emitted photon packets which make up the SED are tagged on
their way to the observer. These tags separate the packets into four
groups. Firstly, packets are separated by source into thermal (disc)
or stellar groups. These groups are then subdivided into those which
reach the observer either directly or after scattering. The resulting
SEDs are discussed in Section \ref{seds}.

@@

Many observational studies use non-spectroscopic data to derive the
pertinent parameters. Therefore in order to examine the practical
effects in the `observational plane' we have used the SEDs to produce
broadband photometric magnitudes, and subsequently colours. Broadband
magnitudes were also derived in a large range of other filter sets not
used explicitly in the analysis within this paper. These magnitudes
are available online\footnote{http://www.astro.ex.ac.uk/research/bd\textunderscore disc}
and are briefly discussed in Appendix \ref{website}. In addition,
monochromatic fluxes have been derived for all filters, and again are
available online and discussed in Appendix \ref{website}.

@@

We have integrated using photon-counting and calibrated using a Vega
spectrum for magnitudes in the optical and near-IR regimes. The filter
bandpasses selected are, the optical system of
\cite{bessell_1998} for \textit{UBVRI} and the \textit{CIT}
system of \cite{elias_1982,stephens_2004} for
\textit{JHK}, with the required shift of $-0.015$\,$\mu$m as prescribed by
\cite{stephens_2004}.

@@

These specific filter sets have been chosen due to their ubiquitous
use and suitability for the derivation of the key stellar parameters
of age, mass and, for populations, disc fractions. Therefore, by
studying the changes of magnitudes (and colours) in these photometric
systems we can test the predicted effects on these derived parameters
caused by changes in the input parameters of our model grid. The
optical magnitudes \textit{VI} are used to explore age related effects
of varying the parameter space. Flux in the \textit{VI} bands is only
minimally affected by accretion flux \citep{gullbring_1998} and
disc thermal flux \citep{hartmann_1998} and, additionally, in a
\textit{V, V$-$I} CMD, the reddening vector lies parallel to the pre-MS,
minimising any age effect of extinction uncertainty, \citep[for a full
discussion see][]{mayne_2008,mayne_2007}. The
near-IR passbands of \textit{JHK} are most often used to derive
stellar masses as for pre-MS objects the reddening vector, in for
instance, a \textit{J, J$-$K} CMD, is almost perpendicular to the
isochrones, minimising the mass effect of extinction uncertainties
\citep[the \textit{CIT} systems was chosen to
match][]{chabrier_2000}. Finally, as is now well documented the
\textit{Spitzer} IRAC passbands provide the best data with which to
unambiguously separate naked and star-disc systems
\citep{luhman_2005a}. In addition, at longer wavelengths, the
MIPS instrument provides sensitivity to disc systems at much larger
radii (or debris discs).

@@

Data for rotation rates, from periodic variability surveys, are widely
available for a range of different age clusters of TTS. However, fewer
studies exist on the rotation rates of BD mass objects. Rotation
period data for $\sigma$ Ori, at an age of $\approx$ 3 Myrs
\citep{mayne_2008}, was studied in \cite{scholz_2004}, where periods
are found over the range 5.78$-$74.4 hours ($\approx$ 0.24$-$3.1 days)
for BD mass objects. \cite{scholz_2005} study rotation period data for
stars in the vicinity of $\epsilon$ Ori, with an assumed age of
$\approx$3 Myrs \citep{osorio_2002} and the ONC at an age of
$\approx$2 Myrs \citep{mayne_2008}. Rotation periods, from photometric
variability, in the range 4.7$-$87.6 hours ($\approx$ 0.2$-$3.65 days)
for BD mass stars are found. \cite{joergens_2003} study the rotational
periods of BD (and very low mass stars) in the Chameleon I
region. This region is $\le$ 1 Myrs old, and the authors find rotation
periods of 2.19, 3.376 and 3.21 days for their BD mass counterparts.

In some cases the periodic variability is irregular and assumed to
come from active accretion hot spots on the BD surface \citep[see][for
a discussion of variability causes]{bouvier_1995,herbst_2007},
indicative of active accretion. All the studies mentioned infer a disc
locking mechanism. Furthermore, \cite{scholz_2004} and
\cite{scholz_2005} find evidence for a mass $\propto$ period
relationship extending into the BD regime. Additionally,
\cite{joergens_2003} propose a shorter lifetime of $\approx$5 Myrs for
BD discs, inferred from a shorter derived disc-locking
timescale. However, as discussed in the previous studies, an imperfect
disc locking mechanisms is also hypothesised as responsible for the
less significant loss in angular momentum out to ages of 10 Myrs, for
BD discs. The data on BD rotation rates, disc presence and disc
locking are summarised and discussed in \cite{herbst_2007}.

@@

As discussed in Section \ref{par_rotation} evidence for irregular
periodic variability has been found in BD stars with detections of
associated stellar discs. This is construed as evidence for accretion
hot spots formed as magnetically channeled material hits the stellar
surface \citep[see discussion
in][]{bouvier_1995,herbst_2007}. The irregularity is
thought to be caused by changes in the magnetospheric structure and
accretion rate \citep{bouvier_1995}. For our model we have
assumed that disc material is disrupted at the co-rotation radius and
channeled onto the star in the form of accretion hot spots with a
characteristic temperature. Therefore, to calculate the characteristic
temperature and the resulting blackbody accretion flux we must adopt
an accretion rate and areal coverage of the accretion stream.

@@

Accretion rates derived for pre-MS stars are of order \logmdot = $-6$
to $-11$ \citep{natta_2006}, with the largest accretion rates found in
so-called FU Orionis type objects. For the more typical accretion
rates \citep[$\dot{M}=$10$^{-11}$ to 10$^{-8.9} M_{\odot}yr^{-1}$, for
TTS,][]{dahm_2009}, several studies have now suggested that the
accretion rate is strongly correlated with the mass of the central
star. This relationship was perhaps first suggested by
\cite{muzerolle_2003} using various accretion diagnostics. Later,
\citep{muzerolle_2005} derived a relationship of approximately
$\dot{M}\propto M_*^2$. Further evidence was put forward by
\cite{natta_2004}, where accretion rates as low as 5$\times$ 10$^{-12}
{\rm M}_{\odot} {\rm yr}^{-1}$ were found for BD stars. More recently,
even lower accretion rates of $\approx$10$^{-13} {\rm M}_{\odot} {\rm
  yr}^{-1}$, have been derived for BDs by \cite{herczeg_2009}.
Further support for a dependence of accretion rate on stellar mass was
apparent in the significantly more homogeneous dataset of
\cite{natta_2006}. \cite{natta_2006} analysed a set of accretion rates
and masses derived for BD stars in $\rho$ Ophiuchi and compared these
results to stars in Taurus. They found that the accretion rate scales
with central object mass into the BD regime, although with significant
scatter.

@@

A further complexity arises when one considers that the density of
material in the disc falls with increasing radius from the star
($\rho(r) \propto r^{-\alpha}$), and that the dust sublimation
temperature is density dependent \citep{pollack_1994}.  Therefore, for
systems where the inner edge has been eroded significantly from the
co-rotation radius the temperature of the inner edge will fall
systematically as the inner dust disc radius increases.

>>>>

There is now strong evidence that as brown dwarfs (BDs) form they pass
through a classical T~Tauri star (CTTS) phase, during which they
possess a flared, dusty circumstellar disc from which they are
actively accreting material
\citep{jayawardhana_2003,mohanty_2004}. The accretion is thought to
proceed via a magnetically-controlled funnel flow mechanism in which
material from a truncated inner disc boundary falls onto the surface
of the star along magnetic field lines
\citep{camenzind_1990,koenigl_1991,muzerolle_2003,mohanty_2008}.

The interpretation of the spectral energy distributions (SEDs) of
pre-main-sequence (pre-MS) stars involves trying to distinguish the
various contributions to the continuum from the hot spots at the base
of the accretion flow, the photospheric flux, and the near-IR flux
from the dusty inner disc. The complexity of the interplay between
these contributions is exacerbated by other geometrical effects such
as the inclination (which changes the projected area of the inner disc
wall visible to the observer) and the outer disc structure
\citep[which may obscure the inner disc for high
inclinations][]{walker_2004,tannirkulam_2007}.

In surveys, particularly those attempting to discern the pre-MS disc
fraction disentangling of the SED components usually takes place using
broad-band photometric measures and cuts in colour-colour or
colour-magnitude space
\citep{luhman_2005b,luhman_2008,gutermuth_2008,monin_2010}.  The
important quantity here is the separation in wavelength between the
emission peaks for the stellar and thermal disc components, as it is
these components which must be isolated. For BD disc (BDD hereafter)
systems it is likely that difficulties detecting the disc will be
exacerbated by the lower temperatures of the photosphere and therefore
the smaller separation in wavelength from the thermal disc emission
component, compared to CTTS systems. Therefore, to derive disc
fractions for BD populations we require detailed comparison models
with which to guide disc candidate selection.

Comparison of accretion rates across the pre-MS mass spectrum has
indicated that the accretion rate is strongly correlated with pre-MS
mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be independent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}. At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photospheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars \citep[although this boundary may well be
spectral type dependent,][]{navascues_2003}. Such objects should show
doppler-broadened profiles, but the line wings will be weak, and the
presence of underlying H$\alpha$ absorption may become dominant. Thus
there may well be a population of low-accretion-rate CTTS which are
current missing from the surveys.

@@

In this section we detail the physical model adopted and assumptions
made (Section \ref{physics}), then explain key elements of the
radiative transfer code (Section \ref{torus}). Then we discuss the
derived values, such as broadband photometric magnitudes and colours
(in Section \ref{derived}). Some internal consistency checks are
given in Appendix \ref{consistency}.

@@

In order to model an accreting BDD system one must first model the
underlying photospheric flux. We have adopted a BD stellar interior
and atmospheric model grid and have then constructed the total
photospheric flux for any input value of stellar age and mass by
interpolating for surface gravity ($\log g$), effective temperature
($T_{\rm eff}$), radius ($R_*/R_{\odot}$) and luminosity
($L_*/L_{\odot}$). These values were then used to interpolate
atmospheric spectra for flux (ergs s$^{-1}$cm$^{-2} {\rm \AA}^{-1}$)
from 1200 to 2$\times$10$^7$ \AA. The spectra were subsequently
resampled onto 200 logarithmically spaced points. Careful inspection
ensured that no spectral features were removed during resampling. The
stellar interior models used for this study are the `DUSTY00' models
of \cite{chabrier_2000} combined with the `AMES-Dusty', atmospheric
models of \cite{chabrier_2000}, which are all available
online\footnote{http://perso.ens-lyon.fr/france.allard/}. For our
$T_{\rm eff}$ range of $\approx$\,3000\,K\,$<T_{\rm eff}<$1600\,K the
AMES-Dusty atmospheres are the most applicable (2700\,K$>T_{\rm
  eff}>$1700\,K). We did try including dynamic application of
atmospheres based on the derived $T_{\rm eff}$, i.e. using AMES-Cond
for $T_{\rm eff}<$1700\,K, but this resulted in large discontinuities
between the model atmospheres and resulting spectra. Additionally, for
the higher temperatures in our range, \cite{martin_2000} have found
the NextGen models to be more applicable for $\sim T_{\rm
  eff}>2300$K. However, as with the lower boundary this only applies
to the edge of our temperature range and we feel it is better to use a
consistent set of atmosphere and interior models across our
grid. Since this only affects stars at the very edge of our
temperature range, i.e. for the oldest and lowest mass objects (for
the AMES-Cond case), we have adopted the AMES-Dusty models throughout.

@@

Finally, a blackbody flux distribution is generated at $T_{\rm acc}$
and added onto the intrinsic stellar photospheric flux. In general,
one would expect this to be an overestimate of the accretion flux, as
for BDs large convective zones are expected on the stellar surface,
and some of the accretion energy may act to further drive these
convective currents, meaning flux is lost. It is worth noting however
that observationally UV excesses are often used to recreate and then
subtract an assumed accretion flux using a blackbody flux curve, which
is essentially the reverse of this method.

@@

In this study we assume that accretion from the central star occurs
along magnetically channelled columns from the inner disc boundary.
For CTTS stars, \cite{bouvier_2007} show that the magnetic truncation
radius ($R_{\rm mag}$) is less than the co-rotation radius ($R_{\rm
  co}$), where the angular Keplerian velocity of the disc is equal to
the surface angular velocity of the central star.  Calculations of the
magnetic truncation radius depend on derivations of the surface
magnetic field \citep{koenigl_1991}. This is currently unavailable for
BDs, due to an increased number of molecular species obscuring
the Zeeman splitting signatures that are normally used to derive
stellar surface magnetic fields. Therefore, for our model grid we have
adopted an initial inner disc radius as the co-rotation radius,
\begin{equation}
{R_{\rm inner}=\left({GM_*\tau^2\over{4\pi ^2}}\right)^{1\over{3}}},
\label{inner_eq}
\end{equation}
where $\tau$ is the stellar rotation period and $R_{\rm inner}$ is the
inner radius. This is effectively adopting a disc-locking mechanism
(without associated angular momentum loss), as for disc-locked stars,
$R_{\rm mag}\approx R_{\rm co}$,
\citep{koenigl_1991,shu_1994}. Therefore, for our
model simulations, this inner edge radius is dependent on, and derived
from, the adopted value of the rotational period for the central star,
as well as being weakly dependent on the stellar mass. As discussed in
Section \ref{intro}, the inner disc can be cleared through a number of
mechanisms, including binarity or giant planet formation,
photoevaporation or photoionisation of the disc and dust grain growth
or settling. For BDD systems where a disc is modeled a treatment of
dust sublimation is included (discussed in Section \ref{dust_edge}).
However, the effects of binarity or giant planet formation are
neglected. Further to the stellar mass and period required prior to
calculation of the inner disc radius, we require a disc mass (in
stellar masses).

@@

We have used the {\sc torus} radiative transfer code which was
originally described by \cite{harries_2000} and was subsequently
updated by \cite{harries_2004} and \cite{kurosawa_2004}. {\sc torus}
uses the method of \cite{lucy_1999} to solve radiative equilibrium on
an AMR mesh. The code also self-consistently solves the equation of
vertical hydrostatic equilibrium and dust sublimation for the disc
\citep[described in][]{tannirkulam_2007}. The {\sc torus} code has
been extensively benchmarked \citep[for radiative transfer in discs
against][]{pinte_2009}.

@@

As our models assume magnetic truncation at the co-rotation radius, we
have generated an inner hole. This also means that the disc will have
an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range \citep{dullemond_2001}. The
circumstellar disc receives direct photospheric radiation and is
strongly heated. This leads to an increased scaleheight, or `puffing
up' of the inner disc rim, and subsequent shadowing of the region
immediately beyond it \citep{dullemond_2001}. The density-dependent
nature of the dust sublimation temperature \citep{pollack_1994} may
also play a role in shaping the inner rim
\citep{isella_2005,tannirkulam_2007}\footnote{Although we recognise
  that the inner hole in protostellar discs may be the result of giant
  planet formation or photoevaporation \citep{dahm_2009,najita_2007},
  these two mechanisms should result in a negligible mass-accretion
  rate \citep{dahm_2009} and therefore would represent a distinct
  evolutionary state of brown dwarf disc (BDD) systems to that of the
  CTTS-like phase we are considering here.}.

@@

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a swift
convergence of the sublimated rim. The dust sublimation proceeds as
follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessitates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<$\,1 at a wavelength of
5500\,\AA. Once a self-consistent sublimated rim has been determined,
the equation of hydrostatic equilibrium is solved throughout the disc,
and the sublimation iterations are restarted (the change in the
density structure from the hydrostatical equilibrium step naturally
feeds back into the shape of the inner rim). Five hydrostatic
equilibrium steps are performed to ensure convergence, although a
stable disc structure is normally found after three such
iterations. For a more sophisticated, and chemically more complex,
study of dust sublimation albeit for higher mass objects see
\cite{kama_2009}.

For our model grid the gas population is assumed to be essentially
static with a zero optical depth.

@@

Once the radiative transfer code was completed simulated SEDs were
generated. These SEDs can be generated for any distance and for any
system inclination. For our models we have set the distance, to 10\,pc
to create absolute flux SEDs and selected ten inclinations equi-spaced
in $\cos i$ (where $i$=0, 27, 39, 48, 56, 64, 71, 77, 84 and
90$^{\circ}$). A further useful feature of the {\sc torus} code is
that the emitted photon packets which contribute to the SED are tagged
on their way to the observer. These tags separate the packets into
four groups. Firstly, packets are separated by source into thermal
(disc) or stellar groups. These groups are then subdivided into those
which reach the observer either directly or after scattering. The
resulting SEDs are discussed in Section \ref{seds}.

@@

Many observational studies use non-spectroscopic data to derive the
pertinent parameters. Therefore in order to examine the practical
effects in the `observational plane' we have used the SEDs to produce
broadband photometric magnitudes and subsequently colours. Broadband
magnitudes were also derived in a large range of other filter sets not
used explicitly in the analysis within this paper. These magnitudes
are available
online\footnote{http://www.astro.ex.ac.uk/research/bd\textunderscore
  disc} and are briefly discussed in Appendix \ref{website}. In
addition, monochromatic fluxes have been derived for all filters, and
again are available online and discussed in Appendix \ref{website}.

@@

We have integrated, the fluxes, using photon-counting and calibrated
using a Vega spectrum for magnitudes in the optical and near-IR
regimes. The filter bandpasses selected are, the optical system of
\cite{bessell_1998} for \textit{UBVRI} and the \textit{CIT} system of
\cite{elias_1982,stephens_2004} for \textit{JHK}, with the required
shift of $-0.015$\,$\mu$m as prescribed by \cite{stephens_2004}.

@@

We have chosen these specific filter due to their ubiquitous use and
suitability for the derivation of the key stellar parameters of age,
mass and, for populations, disc fractions. Therefore, by studying the
changes of magnitudes (and colours) in these photometric systems we
can test the predicted effects on these derived parameters caused by
changes in the input parameters of our model grid. The optical
magnitudes \textit{VI} are used to explore age-related effects of
varying the parameter space. Flux in the \textit{VI} bands is only
minimally affected by accretion flux \citep{gullbring_1998} and disc
thermal flux \citep{hartmann_1998} and, additionally, in a \textit{V,
  V$-$I} CMD, the reddening vector lies parallel to the pre-MS,
minimising any age effect of extinction uncertainty, \citep[for a full
discussion see][]{mayne_2007,mayne_2008}. The near-IR passbands of
\textit{JHK} are most often used to derive stellar masses. This is as
for pre-MS objects the reddening vector, in for instance, a \textit{J,
  J$-$K} CMD, is almost perpendicular to the isochrones. The direction
of this vector acts to minimise the mass effect of extinction
uncertainties \citep[the \textit{CIT} systems was chosen to
match][]{chabrier_2000}. Finally, as is now well documented the
\textit{Spitzer} IRAC passbands provide the best data with which to
unambiguously separate naked and star-disc systems
\citep{luhman_2005a}. In addition, at longer wavelengths, the MIPS
instrument provides sensitivity to disc systems at much larger radii
(or debris discs).

@@

Data for rotation rates, from periodic variability surveys, are widely
available for a range of different age clusters of TTS. However, fewer
studies exist on the rotation rates of BD mass objects. Rotation
period data for $\sigma$ Ori, at an age of $\approx$\,3 Myrs
\citep{mayne_2008}, was studied by \cite{scholz_2004}, where periods
are found over the range 5.78$-$74.4 hours ($\approx$\,0.24$-$3.1
days) for BD mass objects. \cite{scholz_2005} study rotation period
data for stars in the vicinity of $\epsilon$ Ori, with an assumed age
of $\approx$\,3 Myrs \citep{osorio_2002} and the ONC at an age of
$\approx$\,2 Myrs \citep{mayne_2008}. Rotation periods, from
photometric variability, in the range 4.7$-$87.6 hours
($\approx$\,0.2$-$3.65 days) for BD mass stars are
found. \cite{joergens_2003} study the rotational periods of BD (and
very low mass stars) in the Chameleon I region. This region is
$\le$\,1 Myrs old, and the authors find rotation periods of 2.19,
3.376 and 3.21 days for their BD targets.

In some cases the periodic variability is irregular and assumed to
come from active accretion hot spots on the BD surface \citep[see][for
a discussion of variability causes]{bouvier_1995,herbst_2007},
indicative of active accretion. All the studies mentioned infer a disc
locking mechanism. Furthermore, \cite{scholz_2004} and
\cite{scholz_2005} find evidence for a mass $\propto$ period
relationship extending into the BD regime. Additionally,
\cite{joergens_2003} propose a shorter lifetime of $\approx$\,5 Myrs for
BD discs, inferred from a shorter derived disc-locking
timescale. However, as discussed in the previous studies, an imperfect
disc locking mechanisms is also hypothesised as responsible for the
less significant loss in angular momentum out to ages of 10 Myrs, for
BD discs. The data on BD rotation rates, disc presence and disc
locking are summarised and discussed in \cite{herbst_2007}.

@@

As discussed in Section \ref{par_rotation} evidence for irregular
periodic variability has been found in BDs with detections of
associated stellar discs. This is construed as evidence for accretion
hot spots formed as magnetically channeled material hits the stellar
surface \citep[see discussion in][]{bouvier_1995,herbst_2007}. The
irregularity is thought to be caused by changes in the magnetospheric
structure and accretion rate \citep{bouvier_1995}. For our model we
have assumed that disc material is disrupted at the co-rotation radius
and channeled onto the star in the form of accretion hot spots with a
characteristic temperature. Therefore, to calculate the characteristic
temperature and the resulting blackbody accretion flux we must adopt
an accretion rate and areal coverage of the accretion stream.

@@

Accretion rates derived for pre-MS stars are of order \logmdot = $-6$
to $-11$ \citep{natta_2006}, with the largest accretion rates found in
so-called FU Orionis type objects. For the more typical accretion
rates \citep[$\dot{M}=$10$^{-11}$ to 10$^{-8.9} M_{\odot}yr^{-1}$, for
TTS,][]{dahm_2009}, several studies have now suggested that the
accretion rate is strongly correlated with the mass of the central
star. This relationship was first suggested by \cite{muzerolle_2003}
using various accretion diagnostics. Later, \cite{muzerolle_2005}
derived a relationship of approximately $\dot{M}\propto
M_*^2$. Further evidence was put forward by \cite{natta_2004}, where
accretion rates as low as 5$\times$ 10$^{-12} {\rm M}_{\odot} {\rm
  yr}^{-1}$ were found for BDs. More recently, even lower
accretion rates of $\approx$\,10$^{-13} {\rm M}_{\odot} {\rm
  yr}^{-1}$, have been derived for BDs by \cite{herczeg_2009}.
Further support for a dependence of accretion rate on stellar mass was
apparent in the significantly more homogeneous dataset of
\cite{natta_2006}. \cite{natta_2006} analysed a set of accretion rates
and masses derived for BDs in $\rho$ Ophiuchi and compared these
results to stars in Taurus. They found that the accretion rate scales
with central object mass into the BD regime, although with significant
scatter.

@@

A further complexity arises when one considers that the density of
material in the disc falls with increasing radius from the star
($\rho(r) \propto r^{-\alpha}$), and that the dust sublimation
temperature is density dependent \citep{pollack_1994}.  Therefore, for
systems where the inner edge has been eroded significantly from the
co-rotation radius, the temperature of the inner edge will fall
systematically as the inner dust disc radius increases.
